---
title: "36-315 Lecture 10, Spring 2016"
author: "Sam Ventura"
date: "February 15th, 2015"
output: html_document
---

#  Interpretting Histograms and Boxplots

#  Histograms

Histograms divide the range of the variable we're interested in into bins, then count how many observations fall into those bins.

By changing the width of the bins or the number of bins, we are changing the resulting histogram.

Histograms are useful for visualizing the marginal distribution of a single variable.  As we will see later, they are also useful for visualizing the conditional distributions of one variable given another.

Let's work with the Cars93 data again.  


```{r}
library(MASS)
library(ggplot2)
library(gridExtra)
data(Cars93)
#head(Cars93)
```

We'll start by specifying the number of bins in our histogram.

```{r}
base_plot <- ggplot(Cars93)
base_plot + geom_histogram(bins = 15) + aes(x = Fuel.tank.capacity)
```

Try changing the number of bins to 5, 10, 15, 30, etc.  What happens?  

Our interpretation of the distribution changes depending on the number of bins we include:

+  With 5 bins, it appears we have a unimodal distribution (one mode) that is asymetric and has a range of about 6 to 28.
+  With 10 bins, it appears that we have a bimodal distribution (two large modes).  The modes themselves appear to be asymetric.  There also appears to be one **potential** outlier around 27-28.
+  With 15 bins, we see a large spike in the distribution of fuel tank capacity around 16 gallons.  We also see a large mode centered at 16-17 gallons, and some small modes at lower and higher capacities.
+  With 30 bins, it's difficult to say much about the distribution.  We should choose a smaller number of bins to smooth out the distribution a bit more.  Otherwise, we just see a bunch of spikes that don't give us much useful information.
+  On the next lab:  How does `R` / `ggplot` choose the default number of bins / bin width?


Note:  It's best to always say **potential** outlier.  Outliers don't have a universally agreed upon definition in statistics.  Outliers are problem/data-dependent.  

Be careful about removing outliers. Often, they are informative!  We don't want to remove extreme values if the extreme values are the ones that are the most interesting.  

Your data and your graphs tell a story -- outliers are a part of this story, so treat them appropriately.



#  Histograms -- Changing the width of the bins

Now, let's look at the number of passengers.  


```{r}
base_plot + geom_histogram(binwidth = 4) + aes(x = Price)
```

Try specifying different values of the width of the bins.  

+  What happens if we choose a `binwidth` of 1?
+  With a binwidth of 2, we start to see a smoother distribution.
+  With a binwidth of 4, we start to see that the distribution of the car price is skewed right.  There is one large mode in the distribution between $10K and $20K, and the distribution of price tails off as the price increases.  It is rare to see a price below about $8K -- it appears that only one car has a price that low in our dataset.  It is also rare to see a price above $40K -- it appears that only three cars have prices that high.  There is one car with a substantially larger price than the others, costing $60k.

Sometimes, it's nice to find out more information about the extreme-valued cars.

```{r}
Cars93[which(Cars93$Price > 40 | Cars93$Price < 8), ]
```

The Ford Festiva is the cheapest car.  It has a very small engine, is very light, has very low horsepower, and gets great gas mileage!  It does not have airbags though.  Please wear your seatbelt.

The Mercedes-Benz 300E is the most expensive car, at $61.9K.  The other expensive cars are the Cadillac Seville and the Infiniti Q45.


#  Conditional Distributions with Histograms

Let's look at some conditional distributions with histograms.  We can do this with `facet_wrap` and `facet_grid`.

```{r}
base_plot + geom_histogram(binwidth = 4) + aes(x = Price) + facet_wrap(~Type)
```

Each one of the mini-graphs shows the conditional distribution of car price given each particular car type.

Small cars are hte least expensive.  Compact cars are also fairly inexpensive.  Vans have a very tight distribution, with almost all vans priced right around $20K.  Midsize cars have the most variance in their price distribution, ranging from about $12K to $60K.

Is it easier to compare these conditional distributions when the graphs are side-by-side or when they are stacked?

Side-by-side:
```{r, fig.width = 15, fig.height = 3}
base_plot + geom_histogram(binwidth = 4) + aes(x = Price) + facet_grid(.~Type)
```


Stacked:
```{r, fig.width = 3, fig.height = 10}
base_plot + geom_histogram(binwidth = 4) + aes(x = Price) + facet_grid(Type~.)
```


#  Histograms vs. Boxplots

Let's compare histograms and boxplots.  Which gives you more information?

```{r}
my_hist <- base_plot + geom_histogram(bins = 10) + aes(x = Fuel.tank.capacity)
my_box <- base_plot + geom_boxplot() + aes(y = Fuel.tank.capacity, x = factor(1)) + coord_flip()
grid.arrange(my_hist, my_box)
```


With histograms, you can change some parameters (bin-width, number of bins, colors) to get more information.

With boxplots, you just get what you get.

Boxplots aren't bad at giving a quick visual summary of the conditional distirbutions, though:

```{r}
base_plot + geom_boxplot() + aes(x = Type, y = Price)
```

In this graph, you get a very quick measure of the centers and spreads of eacah of the conditional distributions.

On the next lab:  How would you sort these boxes by their medians?

You can flip the x and y axes if you prefer:

```{r}
base_plot + geom_boxplot() + aes(x = Type, y = Price) + coord_flip()
```








